Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-06-18 WholePlotTrajectories.tex
relecture discussion

\Abstract{
Understand the maintenance of forests biodiversity structure,
functioning and dynamics is urgent to anticipate their fate in the
global changing context. Communities diversity structure is assumed to
rely on a constant regime of disturbance and peak at intermediate
intensity. For tropical forests this intermediate disturbance hypothesis
(IDH) remain debated as well as the completeness of forests resilience
in all taxonomic and functoinal facets. To disentangle the ecological
processes driving forests dynamics we examined the trajectories over 30
years after a gradient of logging and thinning disturbance, focusing on
communities taxonomic and functional composition and diversity.
Specifically we analysed the trajectories of communities taxonomic
richness and evennes and functional composition, diversity and
redundancy based on 7 key functional traits assessing species ecology.
We highlighted a decoupling between functional and taxonomic
trajectories, translating a fast recovery of communities functioning
while their taxonomic recovery was slowed by the recruitment of
infrequent species specific to mature forests. The IDH consistently
predicted communities functional trajectories but poorly reflected the
taxonomic ones, blurried by changes in the functional redundancy
entailing competitive exclusion. Although communities consistently
recovered their functioning after 30 years, initially infrequent species
proved long to recover which impacted the taxonomic structure and
functoinal redundancy over several decades. Our results acknowledged the
need of decades-long recovery cycles to restore the resilience of the
community and prevent species loss in the long term.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local climate, support the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The simultaneous
increased demand in forests products and the substantial climatic
changes currently heighten the pressure on remaining forests
\citep{Gibson2011a, Morales-Hidalgo2015}, threatening the maintenance of
communities structure, composition and functioning and their dynamics in
space and time \citep{Anderson-Teixeira2013, Sist2015}.

In tropical forest, ecological communities are constantly re-shaped by
natural disturbance events that change both abiotic environment, through
the fluxes of light, heat and water \citep{Goulamoussene2017}, and
biotic interactions like competition among species \citep{Herault2018}.
The cornerstone of tropical forests ecology is to understand the
mechanisms and the determinants of ecosystems response to disturbance
\citep{White2001, Chazdon2003a}. For now, this has been largely studied
through structural parameters, rapid and convenient to measure, as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. These structural parameters were
thereafter sucessfully modeled, giving important insights into the
maintenance of ecosystems processes and services
\citep{Denslow2000, Blanc2009}. However the response of forests
diversity in tree species remains unclear, albeit it determines the
productivity, stability and functioning of ecosystems
\citep[\citet{Liang2016}]{Tilman2014} and would be most probably
impacted by the changes following disturbance \citep{Baraloto2012a}.

In the short-term disturbance demonstrated negligible or even positive
impacts on communities diversity, which have been formalized by the
intermediate disturbance hypothesis (IDH) stating a maximized species
diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Still, validations of the
IDH remain scarce in the long term and mainly rely on species richness
analyses that give limited or misleading information on forests recovery
and functioning \citep{Martin2015, Chaudhary2016}. More releveant
monitoring would encompass communities composition, crucial for
conservation issues, and complete diversity profiles integrating both
species richness and evenness, to reveal the ecological rules
structuring communities \citep{Magurran1988, Lavorel2002, Bellwood2006}.
Furthermore, the functional approach that accounts for species
biological attributes and role in the ecosystem would be essential to
understand the correlations between ecosystems biodiversity, functioning
and environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. The
functional trait-based approach focused on major traits related to
species ecology and performance was sucessfully adopted
\citep{Diaz2005, Villeger2008a}, for example in highlighting
deterministic processes in tropical rainforests that foster fast growing
species with efficient resources acquisition after disturbance
\citep{Molino2001, Haddad2008, Ruger2009}. Communities then shifted from
a dominance of ``conservative'' slow-growing species dealing with scarce
resources to ``acquisitive'' fast-growing species with rapid and
efficient use of abundant resources
\citep{TerSteege2001, Reich2014, Herault2011}. This shift translated
into consistent trajectories for key functional traits related to
resource acquisition (leaf area, density and chlorophyll content, and
stem specific gravity and bark thickness), tree growth and reproduction
life history traits (seed mass and maximum height)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}.

A combination of taxonomic and functional approaches are essential to
fully assess communities response to disturbance and differences in
their respective trajectories are insightful of underlying mechanisms
\citep{Lohbeck2015, Guariguata2001}. Although taxonomic and functional
diversity are complementary and sometimes decoupled, they are combined
in the measure of communities functional redundancy that quantifies the
amount of shared trait values among species \citep{Carmona2016}.
Functional redundancy is determinant of communities resilience as high
redundancy, like in highly diverse tropical forests
\citep{Bellwood2006}, mitigates the impacts of species removal on
ecosystem functioning \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Grasp all facets of communities response to disturbance comes to examine
the taxonomic and functional trajectories in diversity and composition.
These trajectories would highlight the ecological rules constraining or
not communities dynamics towards the recovery of initial composition,
diversity and functioning. They would therefore clarify the tenants in
the long term of the debated Intermediate Disturbance Hypothesis for
tropical forests, a cruicial point for future adaptive conservation
strategies \citep{Adler2007}. Here we monitored over 30 years the
response of 75 ha of forests plots set up on a gradient of disturbance
intensity, from 10 to 60\% of ecosystem biomass removed. We made use of
a large functional traits database browsing major leaf, stem and seed
functional traits and species maximum height to draw the trajectories
over time of communities taxonomic and functional composition and
diversity. Specifically, we (i) questioned the coupling between
taxonomic and functional response to disturbance and identified the
underlying assembly processes, which allowed to (ii) clarify the
validity of the IDH in the long term for tropical forest and (iii)
question the completeness of communities recovery regarding their
functional redundancy.

@@

The experiment is a network of twelve 6.25ha plots that underwent a
gradient of three logging, thinning and fuelwood cutting treatments
(Table \ref{tab:Tab1}). Disturbance treatments were attributed according
to a randomized plot design with three replicate blocks of four plots.
The disturbance corresponds to averages of 10 trees removed per hectare
with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1),
32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees/ha above
40 cm DBH for treatment 3 (T3). Treatments T2 and T3 besides included
the thinning of trees by poison girdling \citep{Schmitt1989, Blanc2009}.
The disturbance intensity was measured as the percentage of aboveground
biomass (\%AGB) lost between the first inventory in 1984 and five years
after disturbance (ref to be found) measured with the BIOMASS R package
\citep{Biomass2018}.

@@

Both communities taxonomic and functional diversity and composition
proved resilient, following similar humped shaped trajectories returning
towards initial values. The resilience of communities functional
structure, that is the most direct link between biodiversity and
ecosystem functioning \citep{Diaz2005}, meant a consistent recovery of
ecosystem processes in the long term \citep{Guariguata2001}. The
resilience of communities taxonomy meant the maintenance of their
initial differences in composition and diversity, suggesting the
existence of multiple stable equilibria as assumed for highly diverse
and productive ecosystems \citep{Chase2003}, and the dependency of
recovery trajectories on the initial composition to be recovered that
would the the asymptot of the humped shaped dynamics
\citep{Hubbell1999, Molino2001, Anderson2007, Baraloto2012a}.

Although both communities taxonomic and functional characteristics
proved resilient and followed similar humped shaped trajectories, the
taxonomic recovery systematically lagged behind the functional one. The
decoupling between functional and taxonomic dynamics has already been
observed for grasslands \citep{Tilman1997, Mouillot2011} and more
recently for tropical forests \citep{Lohbeck2015, Guariguata2001}.
According to the ``vegetation quantity effect'' \citep{Grime1998}
functional trajectories rely on the pool of dominant species, which
diversity and evenness increased after disturbance but rapidly recovered
their initial functional structure. Although the pool of dominant
species recovered quickly, communities evenness remained high so
infrequent species still missed to the taxonomic recovery. Taxonomic
recovery was then all the more longer that unrecovered species would be
functionally redundant and probably underwent selective competition
processes.

@@

The functional and taxonomic decoupling and the disprove of the IDh for
communities taxonomic structure were explained were explained by
persisting changes in communities functoinal redundancy that implied
communities incomplete resilience
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Functional redundancy mitigates the loss of species on communities
functional structure \citep{Carmona2016}, and therefore reduces the
impact on disturbance. The functional redundancy itself however might be
reduced or re-organized within the functional space and alter
communities resilience. Although the functional redundancy considered in
the whole communities functional space did not follow consistent
trajectories after disturbance, the redundancy restricted to the
functional space of the initial community clearly followed humped shaped
trajectories. This restricted redundancy first decreased, all the more
so that the disturbance was intense, before returning towards the
initial value but this had not recovered for any disturbed community
after 30 years. After disturbance communities differently occupied the
functional space following traits trajectories and the shifts in
functional composition, before recovering the initial functional space
and its high redundancy. This recovery phase was slow as it depend on
infrequent, functionally redundant species specific to old-growth
forests that rely upon different ecological processes, stochastic or
deterministic competitive pressures limiting species functional
similarity \citep{Chave2004, Mayfield2010}.

The unachieved recovery of the functional redundancy meant a decreased
resilience of pre-disturbance communities and an increased of
disturbance-specific communities (corresponding to lower richness and
higher dominance of pioneer species). After disturbance, chances were
higher to observe long lasting or self-maintained compositional changes
in favor of disturbance resistant species, lianas or epiphytes
\citep{Haddad2008, Burslem2000, Martin2013}. The recovery remained
incomplete until the recovery of the initial functional redundancy and
resilience, and this specifically threatened the maintenance of species
contingent to undisturbed forests and increased the risk of loosing
cornerstone species and having unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}.

@@

Our study defined the decoupled functional and taxonomic trajectories of
tropical rainforests after disturbancethat demonstrated a rapid
functional recovery but a slower and more variable taxonomic one.
Consistently with the IDH, functional trajectories were driven by
deterministic processes favoring pioneers and light-demanding species
after disturbance. The following functional shifts entailed a
re-organization of communities functional redundancy in the functional
space that proved long to recover and explained the more variable and
longer taxonomic trajectories as shade tolerant species of old-growth
forests faced functional redundancy and limiting similarity processes.
The resilience of tropical forests then proved long in the face of quite
intense disturbance and did not preclude the settling of a persistent
disturbanc-specific community \citep{Gourlet-Fleury2005}. Within the
mechanisms of this resilience the recruitment processes proved central,
in accordance with the IDH, and would deserve closer focus to specify
their ecological drivers. The disturbance range however stayed within
the spectrum of selective logging, with a forest cover remaining all
along the experiment, and the response mechanisms would probably be much
different after harder disturbance.

>>>>

\Abstract{
Understand the maintenance of forests biodiversity structure,
functioning and dynamics is urgent to anticipate their fate in the
global changing context. Communities diversity are assumed to be shaped
by constant regime of disturbance yielding a diversity peak at
intermediate intensity, but for tropical forests this intermediate
disturbance hypothesis (IDH) remains debated as well as the completeness
their resilience in all taxonomic and functional facets after
disturbance. To disentangle the ecological processes driving communities
response to disturbance we analysed the taxonomic and functional
trajectories following a logging and thinning disturbance gradient.
Specifically we examined over 30 years the trajectories of communities
taxonomic richness and evennes and functional composition, diversity and
redundancy based on 7 leaf, stem and life history traits assessing
species ecology. Trajectories highlighted communities taxonomic
resilience, maintaining their initial differences in composition, and
functional resilience through fast functional trajectory common to all
communities in the functional space. The IDH consistently predicted
communities functional trajectories according to the disturbance
intensity, but poorly represented their taxonomic trajectories, blurried
by the alteration of the functional redundancy and the recruitment of
infrequent, shade tolerant species slowed by competitive exclusion
processes. Although communities functioning had recovered after 30 years
their taxonomic diversity and composition in infrequent species and
their functional redundancy remained altered. these results acknowledged
the need of decades-long recovery cycles to ensure communities complete
resilience, and questioned communities maintenance after repeated
disturbance.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local climate, and the carbon, water and nutrient cycles,
and ensure cultural and human well-being. The growing demand in forests
products together with current climatic changes increases the pressure
on remaining forests \citep{Gibson2011a, Morales-Hidalgo2015}and
threatens the maintenance and dynamics in space and time of communities
structure, composition and functioning
\citep{Anderson-Teixeira2013, Sist2015}.

In tropical forest, ecological communities are constantly re-shaped by
natural disturbance events that change both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions like competition among species
\citep{Herault2018}. The cornerstone of tropical forests ecology is to
understand the processes and the determinants of ecosystems response to
disturbance \citep{White2001, Chazdon2003a}. For now, this has been
largely studied through structural parameters, rapid and convenient to
measure, as aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. These structural parameters were
thereafter sucessfully modeled, giving important insights into the
maintenance of ecosystems processes and services
\citep{Denslow2000, Blanc2009}. However the response of forests
diversity in tree species remains unclear, albeit it determines the
productivity, stability and functioning of ecosystems
\citep[\citet{Liang2016}]{Tilman2014} and would be most probably
impacted by the changes induced by disturbance \citep{Baraloto2012a}.

In the short-term disturbance demonstrated negligible or even positive
impacts on communities diversity, which have been formalized by the
intermediate disturbance hypothesis (IDH) stating a maximized species
diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Still, validations of the
IDH remain scarce in the long term and mainly rely on the analysis of
species richness that gives limited or misleading information on forests
recovery and functioning \citep{Martin2015, Chaudhary2016}. More
complete analysis would besides encompass communities explicit
composition that is crucial for conservation issues, and communities
evenness to reveal the ecological rules structuring communities
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, a
functional approach accoutning for species biological attributes and
role in the ecosystem directly links ecosystems biodiversity,
functioning and environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. The
functional trait-based approach focused on major traits related to
species ecology and performance was sucessfully adopted
\citep{Diaz2005, Villeger2008a}, for example to reveal in tropical
rainforests the deterministic processes that foster after disturbance
the fast growing species with efficient resources acquisition
\citep{Molino2001, Haddad2008, Ruger2009}. This basic deterministic
process entails a functional shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources to ``acquisitive''
fast-growing species with rapid and efficient use of abundant resources
\citep{TerSteege2001, Reich2014, Herault2011}. This shift is translated
into consistent trajectories of key functional traits related to
resource acquisition (leaf area, density and chlorophyll content, and
stem specific gravity and bark thickness), tree growth and reproduction
life history traits (seed mass and maximum height)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Besides
communities functional composition and diversity, a key aspect of
communties response to disturbance is the changes in functional
redundancy that quantifies the amount of shared trait values among
species \citep{Carmona2016}. High functional redundancy, as in the very
diverse tropical forests \citep{Bellwood2006}, mitigates the impacts of
species removal on ecosystem functioning and determines communities
resilience \citep{Trenbath1999, Elmqvist2003, Diaz2005}. High functional
redundancy also increases the functional overlap among species and would
foster neutral stochastic processes \citep{Gravel2006}.

To grasp all facets of communities response to disturbance we examined
here the taxonomic and functional trajectories in diversity and
composition \citep{Lohbeck2015, Guariguata2001}. These trajectories
would highlight the recovery of communities initial composition,
diversity and functioning and the underlying ecological processes. They
would clarify the tenants in the long term of the Intermediate
Disturbance Hypothesis debated in tropical forests, and provide
indications for future adaptive conservation strategies
\citep{Adler2007}. We monitored over 30 years the response of 75 ha of
neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem biomass removed. We made use of a large
functional traits database browsing major leaf, stem and seed functional
traits and species maximum height to draw the trajectories over time of
communities taxonomic and functional composition and diversity.
Specifically, we (i) questioned the coupling between taxonomic and
functional response to disturbance and identified the underlying
assembly processes, (ii) clarified the validity of the IDH in the long
term for tropical forest and (iii) questioned the completeness of
communities recovery regarding communities functional redundancy.

@@

The experiment is a network of twelve 6.25ha plots that underwent a
gradient of three logging, thinning and fuelwood cutting treatments
(Table \ref{tab:Tab1}) attributed according to a randomized plot design
with three replicate blocks of four plots. The disturbance corresponds
to averages of 10 trees removed per hectare with a diameter at 1.3 m
height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm
DBH for treatment 2 (T2) and 40 trees/ha above 40 cm DBH for treatment 3
(T3). Treatments T2 and T3 besides included the thinning of trees by
poison girdling \citep{Schmitt1989, Blanc2009}. The disturbance
intensity was measured as the percentage of aboveground biomass (\%AGB)
lost between the first inventory in 1984 and five years after
disturbance (ref to be found) measured with the BIOMASS R package
\citep{Biomass2018}.

@@

Both communities taxonomic and functional diversities and composition
proved resilient, following similar humped shaped trajectories returning
towards initial values. Communities functional structure is the most
direct link between biodiversity and ecosystem functioning
\citep{Diaz2005} and its resilience meant the recovery of ecosystem
processes in the long term \citep{Guariguata2001}. The resilience of
communities taxonomy in turn meant the maintenance of their initial
differences in species composition and diversity. The maintenance of
initial differences suggested the existence of multiple stable
equilibria as assumed for highly diverse and productive ecosystems
\citep{Chase2003} and the dependency of recovery trajectories on the
initial composition to be recovered
\citep{Hubbell1999, Molino2001, Anderson2007, Baraloto2012a}.

Although communities taxonomic and functional trajectories were similar
the taxonomic recovery lagged behind the functional one, confirming a
decoupling between functional and taxonomic dynamics already observed
for grasslands \citep{Tilman1997, Mouillot2011} and tropical forests
\citep{Lohbeck2015, Guariguata2001}. According to the ``vegetation
quantity effect'' \citep{Grime1998} the functional trajectories rely on
the pool of dominant species, which diversity and evenness increased and
rapidly recovered after disturbance. At the same time communities
taxonomic evenness remained high, revealing the unachieved recovery of
infrequent species.

@@

The functional and taxonomic decoupling and the disprove of the IDH for
communities taxonomic structure were explained by persisting rearranging
of communities functional redundancy that implied an incomplete
resilience of communities \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Although the functional redundancy considered in the functional space of
the whole communities did not display consistent trajectories after
disturbance, the redundancy restricted to the functional space of the
initial community clearly followed humped shaped trajectories. After
disturbance communities differently occupied the functional space,
following the observed shifts in traits trajectories and functional
composition, and the redundancy in the initial functional space
decreased according to the disturbance intensity. After 30 years the
recovery of the initial redundancy structure was consistent but remained
unachieved. It depended on infrequent species specific to old-growth
forests that were functionally and therefore underwent competitive
exclusion limiting functional similarity
\citep{Chave2004, Mayfield2010}. The persistent alteration of functional
redundancy meant a lower resilience of pre-disturbance communities and
an higher resilience disturbance-specific communities (corresponding to
lower richness and higher dominance of pioneer species) so chances to
observe self-maintained compositional changes in favor of disturbance
resistant species, lianas or epiphytes were increased after disturbance
\citep{Haddad2008, Burslem2000, Martin2013}. Communities resilience
depended on the functional redundancy structure that proved long to
recover. This decades-long recovery specifically increased the risks of
loosing cornerstone specie, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}.

@@

Our study defined the decoupled functional and taxonomic trajectories of
tropical rainforests after disturbance demonstrating a rapid functional
recovery but a slower and more variable taxonomic one. Consistently with
the IDH, functional trajectories were driven after disturbance by
deterministic processes favoring pioneers and light-demanding species.
The following functional shifts entailed a re-organization of
communities functional redundancy that proved long to recover and
explained the more variable and longer taxonomic trajectories as shade
tolerant species of old-growth forests faced functional redundancy and
limiting similarity processes. The resilience of tropical forests then
proved long in the face of quite intense disturbance and did not
preclude the settling of a persistent disturbance-specific community
\citep{Gourlet-Fleury2005}. As the trajectories highlighted the
recruitment processes proved central for communities response to
disturbance and closer focus on demographical drivers of communities
response would clarify the fate of the future forests. The disturbance
range however stayed within the spectrum of selective logging, with a
forest cover remaining all along the experiment, and the response
mechanisms would probably be much different after harder disturbance.
